The Evolution of Phenotypically Invariant Genetic Networks

### 2021-04-14 review-responses.tex

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are pleased to submit a revision of our manuscript, ``{System drift and speciation},'' for your review. 
We apologize for the delayed resubmission but hope that you will find the
revised manuscript significantly improved.

We thank you and the two reviewers for providing constructive suggestions 
and feedback that has helped to improve the manuscript. We have followed 
the suggestions closely and hope to have fully addressed all the concerns and 
suggestions as detailed in our pont-by-point response below.

In our original submission, we studied the effect of gene regulatory network evolution on speciation,
first by applying linear system theory to characterize the large space of phenotypically equivalent network organizations,
and second by using quantitative genetics to show that neutral genetic drift on this network space can lead to speciation over plausible timescales.

We agree with the reviewers that the combination of these two approaches, while interesting,
was not seamless and that the quantitative genetics section required further development, 
and thus will follow the recommendation of both the Editor and Associate Editor, to break the manuscript into
two papers. One paper will focus on quantitative genetics, and the other -- 
the focus of this present revision -- will focus on system theory. 
A revised manuscript focusing on quantitative genetics will be submitted at a later date.

We hope that by focusing this paper system theory, we have clarified our results, which include a mathematical description of phenotypically equivalent network space 
and its implications for speciation under neutral genetic drift. These results suggest that the space of phenotypically equivalent network organizations is substantial,
and often not reproductively compatible, leaving many oportunities for independelty evolving populations to become reproductively incompatible.
The linear system framework applied in this manuscript explicitly describes possible moleuclar pathways and naturally predicts Haldane's rule.

We would like to note that the previous version of this paper as a preprint has since recieved 10 citations as well as attracted
substantial positive attention as a preprint (e.g., see tweets at
\url{https://www.biorxiv.org/content/10.1101/231209v2}). Furthermore, this paper has been downloaded more than 2,229 times \url{https://rxivist.org/papers/24911},
making it the 156$^{\text{th}}$ most downloaded evolutionary biology paper available on the bioRxiv, as of this writing.

### 2021-04-27 kalman_walk_paper.tex
edits through rev 1

It is an overarching goal of many biological subdisciplines 
to attain a general understanding of the function and evolution of the 
complex molecular machinery that translates an organism's genome 
into the characteristics on which natural selection acts,
the phenotype.
For example, 
there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
Mathematical models of both particular regulatory networks
and the evolution of such systems in general
can provide guidance where intuition fails,
and thus has the potential to discover general principles in the organization of biological systems 
as well as provide concrete numerical predictions \citep{servedio2014not}.
There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness}
and empirically inspired
\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.

At all levels of biological organization,
the problems that biological systems have evolved to solve
often do not have single solutions --
systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
Examples can be found across nearly all levels of biological organization from the level of the genetic code itself % \citep{crick1963recent}
all the way up to the convergent evolution of adaptive traits.
In many cases, these functionally equivalent structures
can be explored through small, local changes to the structure that leave the function unchanged.
For instance, there are ``neutral networks''
of nucleic acid sequences that produce the same RNA secondary structure \citep{gruner1996analysis}\revpoint{1}{5},
amino acid sequences that fold similarly \citep{babajide1997neutral},
or proteins with equivalent thermodynamic stability \citep{hart2014thermodynamic}.
Further examples are found in the vast space of functionally equivalent
potential regulatory sequences \citep{hare2008sepsid}, 
in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription} 
and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{true2001developmental}. 
\revpoint{1}{1} \revpoint{1}{3}
% \jss{Other ideas instead of ``degeneracy'' or ``isofunctional structural plasticity'': non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
% \plr{'structural plasticity' means something specific in neuroscience}

This capacity for isofunctional yet distinct mechanisms, sometimes called \emph{degeneracy},
is a consequence of a many-to-one mapping between a system's structure and function,
a concept that has been explored in many fields beyond biology.
For instance, in many contexts mathematical models
are fundamentally \emph{nonidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, despite access to complete and perfect data \citep[e.g.,][]{bellman1970structural, grewal1976identifiability, walter1984structural}. 
Models with different parameter schemes, or even different mechanics 
can make equally accurate predictions,
but still not actually reflect the internal dynamics of the system being modeled.
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a ``black box'',
even if all possible input and output experiments are performed 
\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
The inherent nonidentifiability of chemical reaction networks is sometimes
referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
In computer science, this has been framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
Finally,
the field of \emph{inverse problems} studies those cases in which,
despite the existence of a theoretical one-to-one mapping between a model and behavior,
  tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}. \revpoint{1}{2}

It has been argued that the ability to modify structure without affecting function
is necessary for natural selection \citep{edelman2001degeneracy}, 
as it may function as a mechanism for biological robustness and evolvability, or manifest as 
\emph{canalization} \citep{whitacre2010degeneracy}.
It may even contribute to the formation of new species \citep{gavrilets2014models}. 
Redundancy of the genetic code, for instance,
can make sequences more fault-tolerant to mutations \citep{sonneborn1965degeneracy},
and robustness to modification of genetic networks can allow adaptation 
without passing through a fitness valley \citep{wagner2008robustness}.

In this paper we use results on mathematical nonidentifiability from
linear systems theory to study how gene regulatory networks can be modified
while retaining the same function, and the possible implications for speciation.
% Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
If system architectures are not functionally unique,
can this open up neutral evolutionary paths,
and do these paths manifest as \emph{developmental system drift} \citep{true2001developmental}?
Is this fast enough to contribute meaningfully to speciation?
To do this, we describe results on linear dynamical systems which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes,
and use quantitative genetics theory to estimate the speed at which system drift can lead to reproductive incompatibility and hence speciation.
In this model,
a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}\revpoint{2}{12}.
% Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.

Besides those already cited, many other authors have observed
% It is not a new observation
that there is often more than one way to do the same thing, 
and that speciation can be the result of (nearly) neutral processes.
Indeed, \citet{bateson1909heredity} first proposed that what today we call a
Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. \revpoint{1}{10}
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. \revpoint{1}{9}
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
Our main aim is to provide a concrete framework 
that can provide natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

\emph{Note on implementation:}
The \emph{reachable subspace},
which we denote by $\reachable$,
is defined to be the closure of $\spn(B)$ under applying $A$
(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$), \revpoint{AE}{6}
% Analogously to this, we define
and the \emph{unobservable subspace}, 
%$\mathcal{O}$, is the closure of $\spn(C^T)$ under applying $A$.
denoted $\unobservable$, is the largest $A$-invariant subspace
contained in the null space of $C$.
% and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
The four subspaces, $\rno$, $\ro$, $\nrno$, and $\nro$
are defined from these by intersections and orthogonal complements --
$\ro$ refers to the both \emph{reachable and observable} subspace,
while $\nrno$ refers to the \emph{unreachable and unobservable} subspace,
and similarly for $\nro$ and $\rno$.

@@

For the remainder of the paper, we interpret $\allS$ as the neutral set in the fitness landscape, 
along which a large population will drift under environmental and selective stasis. 
Even if the phenotype is constrained and remains constant through evolutionary time, 
the molecular mechanism underpinning it is not constrained and likely will not be conserved.

@@

Exactly how the coefficients 
(i.e., entries of $A$, $B$ and $C$)
of a haploid system inherited by an offspring from her diploid parent
are determined by the parent's two systems
depends on the genetic basis of any variation in the coefficients.
Thanks to the randomness of meiotic segregation,
the result is random to the extent that each parent is heterozygous
for alleles that affect the coefficients.
Since the $i^\text{th}$ row of $A$ summarizes how each gene regulates gene $i$,
and hence is determined by the promoter region of gene $i$,
the elements of a row of $A$ tend to be inherited together,
which will create covariance between entries of the same row.
It is, however, a quite general observation that the variation seen among recombinant systems
is proportional to the difference between the two parental systems.
% This is certainly true if each coefficient is determined by a single nonrecombining locus,
% so that each coefficient in the system produced by meiosis is an independent random choice
% between the two parental coefficients.
% On the other hand, if a coefficient is determined by additive contributions of independently segregating loci,
% then the law of total variance applied after conditioning on the parental origin of each allele
% implies that the variance is equal to one-quarter of the squared difference between parental coefficients.

@@

\begin{figure}[H]
\centering
\includegraphics{figures/conceptual_fig}
\caption{
    \label{fig:conceptual_fig}
    A conceptual figure of the fitness consequences of hybridization:
    axes represent system coefficients (i.e., entries of $A$);
    the line of optimal system coefficients is down in black;
    solid lines give phenotypic distances to the optimum.
    A pair of parental populations are shown in black, along the optimum;
    a hypothetical population of $F_1$s are shown in red,
    and the distribution of $F_2$s is shown in purple.
    The two figures differ in the genetic basis, and hence,
    the distribution of $F_2$ phenotypes:
    \textbf{(left)}
    $F_2$s compose all four mixed homozygotes if variation at both
    traits has a simple, one-locus genetic basis in both popualtions; and
    \textbf{(right)}
    $F_2$ show a much wider distribution of phenotypes
    if the genetic basis of variation in each population is polygenic.
}
\end{figure}

\subsection*{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
How quickly do hybrid phenotypes break down as genetic distance between parents increases?
We will quantify how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative weighting function, \revpoint{2}{3}
$h_0(t)$ is the \emph{optimal} impulse response function
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
\label{eqn:distance}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
In practice, we take $\rho(t) = \exp(-t/4\pi)$,
so that fitness is determined by the dynamics of the system over a few multiples of $2 \pi$,
but not longer. \revpoint{AE}{8}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ 
is explicitly worked out in Appendix \ref{apx:H_calc}, and \revpoint{2}{4}
shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry
of the optimal set $\allS$ relative to segregating genetic variation.
% in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set) with respect to the coordinates.

\begin{example}[Hybrid incompatibility: misregulation due to system drift] \label{ex:hybrid_osc}
Offspring of two equivalent systems from Example \ref{ex:all_osc}
can easily fail to oscillate.
For instance, the $F_1$ offspring between homozygous parents at $\tau=0$ and $\tau=-2$
has phenotype $\phi_{F_1}(t) = e^t$, rather than $\phi(t) = \sin t + \cos t$.
However, the coefficients of these two parental systems differ substantially,
probably more than would be observed between diverging populations.
In figure \ref{fig:hybs} we compare the phenotypes for $F_1$ and $F_2$ hybrids between more similar parents,
and see increasingly divergent phenotypes as the difference between the parental systems increases.
(In this example, the coefficients of $A(\epsilon)$ differ from those of $A(0)$ by an average factor of $1+\epsilon/2$;
such small differences could plausibly be caused by changes to promoter sequences.)
This divergence is quantified in Figure \ref{fig:osc_incompat},
which shows that mean distance to optimum phenotype of the $F_1$ and $F_2$ hybrid offspring between $A(0)$ and $A(\epsilon)$
increases with $\epsilon^2$ and $\epsilon$, respectively.

\begin{figure}[H]
  \centering
 % \begin{tabular}{ccc}
 %   \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_hundreth_tau0} &
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_hundrethtau0} \\
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_tenth_tau0.pdf} &
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_tenthtau0.pdf} \\
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_half_tau0.pdf} &
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_halftau0.pdf}
 % \end{tabular}
  \includegraphics[scale=0.5]{examples/f1f2_newplot}
  \caption{
    \textbf{(left)} Phenotypes of $F_1$ hybrids between a homozygous $A(0)$ parent and, 
    top-to-bottom, homozygous $A(\sfrac{1}{100})$, $A(\sfrac{1}{10})$, and $A(\sfrac{1}{2})$ parents;
    parental coefficients differ by around 0.5\%, 5\%, and 25\% respectively.
    Parental phenotypes ($\sin t + \cos t$) are shown in solid black, and hybrid phenotypes in dashed blue.
    \textbf{(right)} Phenotypes of all $3^4 = 81$ possible $F_2$ hybrids between the same set of parents,
    with parental phenotype again in black.
    %% PLR: There are at most 128 possible F2s, not 256; but many are equivalent; 
    %%   it only matters if the offspring is homo/het/homozygous, so there are 3^4 = 81 possible expression patterns.
    Different colored lines correspond to different $F_2$ hybrids;
    note that some completely fail to oscillate.
  } \label{fig:hybs}
\end{figure}

  \begin{figure}[H]
    \centering
    \includegraphics[width=0.5\textwidth]{examples/F2_vs_F1_divergence_tau0}
    \caption{Mean hybrid phenotypic distance from optimum computed with equation \eqref{eqn:distance}, using $\rho(t) = \exp(-t/4\pi)$
    for $F_1$ (black) and $F_2$ (blue) hybrids between $A(0)$ and $A(\epsilon)$ parent oscillators. Genetic distance is computed as 
    $\left(\sum_{ij} (A_{ij}(0) - A_{ij}(\epsilon))^2\right)^{1/2}$.
    }
    \label{fig:osc_incompat}
  \end{figure}
\end{example}

>>>>

It is an overarching goal of many biological subdisciplines 
to attain a general understanding of the function and evolution of the 
complex molecular machinery that translates an organism's genome 
into the characteristics on which natural selection acts,
the phenotype.
For example, 
there is a growing body of data on the evolutionary histories and molecular characterizations of particular gene regulatory networks
\citep{jaeger2011gap, davidson2006gene, israel2016comparative}, 
as well as thoughtful verbal and conceptual models \citep{true2001developmental, weiss2000phenogenetic, edelman2001degeneracy, pavlicev2012model}. 
Mathematical models of both particular regulatory networks
and the evolution of such systems in general
can provide guidance where intuition fails,
and thus have the potential to discover general principles in the organization of biological systems 
as well as provide concrete numerical predictions \citep{servedio2014not}.
There is a substantial amount of work studying the evolution of gene regulatory networks, in frameworks 
both abstract \citep{wagner1994evolution, wagner1996does, siegal2002waddington, bergman2003evolutionary, draghi2015robustness}
and empirically inspired
\citep{mjolsness1991connectionist, jaeger2004dynamic, vitaly1, crombach2016gap, wotton2015quantitative, chertkova2017insilico}.

At all levels of biological organization,
the problems that biological systems have evolved to solve
often do not have single solutions --
systems can be structurally different yet remain functionally equivalent \citep{edelman2001degeneracy}. 
Examples can be found across nearly all levels of biological organization from the level of the genetic code itself % \citep{crick1963recent}
all the way up to the convergent evolution of adaptive traits.
In many cases, these functionally equivalent structures
can be explored through small, local changes to the structure that leave the function unchanged.
For instance, there are ``neutral networks''
of nucleic acid sequences that produce the same RNA secondary structure \citep{gruner1996analysis}\revpoint{1}{5},
amino acid sequences that fold similarly \citep{babajide1997neutral},
or proteins with equivalent thermodynamic stability \citep{hart2014thermodynamic}.
Further examples are found in the vast space of functionally equivalent
potential regulatory sequences \citep{hare2008sepsid}, 
in the logic of transcriptional \citep{tsong2006evolution, matsui2015regulatory, dalal2016transcriptional, dalal2017transcription,jimenez2017spectrum} 
and neural circuits \citep{trojanowski2014neural}, and in developmental systems \citep{vondassow2000segment,true2001developmental}. 
\revpoint{1}{1} \revpoint{1}{3}
% \jss{Other ideas instead of ``degeneracy'' or ``isofunctional structural plasticity'': non-invertibility, isofunctional plasticity, structural plasticity, phenotypic equivalency, phenotypic invariance, functional invariance, isofunctionality, functional equivalency, genotypic plasticity, conjugacy, functional symmetry, isologous, isomorphic.}
% \plr{'structural plasticity' means something specific in neuroscience}

This capacity for isofunctional yet distinct mechanisms, sometimes called \emph{degeneracy},
is a consequence of a many-to-one mapping between a system's structure and function,
a concept that has been explored in many fields beyond biology.
For instance, in many contexts mathematical models
are fundamentally \emph{nonidentifiable} and/or \emph{indistinguishable} -- \revpoint{1}{2}
meaning that there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, despite access to complete and perfect data \citep[e.g.,][]{bellman1970structural, grewal1976identifiability, walter1984structural}. 
Models with different parameter schemes, or even different mechanics 
can make equally accurate predictions,
but still not actually reflect the internal dynamics of the system being modeled.
In control theory, where electrical circuits and mechanical systems are often the focus, 
it is understood that there can be an infinite number of ``realizations'', 
or ways to reverse engineer the dynamics of a ``black box'',
even if all possible input and output experiments are performed 
\citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. 
The inherent nonidentifiability of chemical reaction networks is sometimes
referred to as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}.
In computer science, this has been framed as the relationship among processes that \emph{simulate} one another \citep{van2004equivalence}.
Finally,
the field of \emph{inverse problems} studies those cases in which,
despite the existence of a theoretical one-to-one mapping between a model and behavior,
tiny amounts of noise make inference problems nonidentifiable in practice \citep{petrov2005well}.

It has been argued that the ability to modify structure without affecting function
is necessary for natural selection \citep{edelman2001degeneracy}, 
as it may function as a mechanism for biological robustness and evolvability
\citep[reviewed in][]{devisser2003perspective}, or manifest as 
\emph{canalization} \citep{whitacre2010degeneracy}.
It may even contribute to the formation of new species \citep{gavrilets2014models}. 
Redundancy of the genetic code, for instance,
can make sequences more fault-tolerant to mutations \citep{sonneborn1965degeneracy},
and robustness to modification of genetic networks can allow adaptation 
without passing through a fitness valley \citep{wagner2008robustness}.

In this paper we use results on mathematical nonidentifiability from
linear systems theory to study how gene regulatory networks can be modified
while retaining the same function, and the possible implications for speciation.
% Specifically, we would like to understand the degree of which gene regulatory network architectures are functionally equivalent and how this relates to network size and efficiency.
If system architectures are not functionally unique,
can this open up neutral evolutionary paths,
and do species explore these paths
through the process termed \emph{developmental system drift} \citep[by][]{true2001developmental}?
Is this fast enough to contribute meaningfully to speciation?
To do this, we describe results on linear dynamical systems which give
an analytical description of the set of 
all linear gene network architectures that yield identical phenotypes,
and use quantitative genetics theory to estimate the speed at which system drift can lead to reproductive incompatibility and hence speciation.
In this model,
a population diffuses along the neutral ridges of a high-dimensional space of possible system parameters,
in a similar vein as \emph{holey landscape} models \citep{gavrilets1997evolution, yamaguchi2013first, pina2019does}\revpoint{2}{12}.
% Our model is developed in more detail in a companion paper \citep{schiffman2020genetic}.

% It is not a new observation
The field of population genetics has also explored the consequences of the fact
that there is often more than one way to do the same thing, 
and observed that speciation might be the result of changes that are themselves neutral.
Indeed, \citet{bateson1909heredity} first proposed that what today we call a
Bateson-Dobzhansky-Muller incompatibility would likely arise through neutral changes. \revpoint{1}{10}
The potential for speciation has been analyzed in models of 
traits under stabilizing selection determined additively by alleles at many loci 
\citep{wright1935evolution,barton1986maintenance,barton1989divergence,barton2001role},
in related fitness landscape models \citep{fraisse2016genetics},
and for pairs of traits that must match but whose value is unconstrained \citep{sved1981twosex}.
% Recent work has shown that Fisher's geometric model of stabilizing selection
% predicts many general patterns from the speciation literature \citep{fraisse2016genetics}.
It has also been shown that population structure
can allow long-term stable coexistence of incompatible genotypes encoding identical phenotypes \citep{phillips1996maintenance}. \revpoint{1}{9}
However, previous simulations of system drift in regulatory sequences \citep{tulchinsky2014hybrid}
and a regulatory cascade \citep{porter2002speciation}
found rapid speciation under directional selection
but only equivocal support for speciation under models of purely neutral drift.
%Additionally, \citet{fierst} observed speciation due to an increase of segregation variance under balancing selection, in models of weak to moderate, but not strong epistasis.
The rate at which hybrid incompatibility accumulates due to genetic drift creating
segregation variance between isolated populations is fairly well understood
\citep{slatkin1994segregation,rosas2010cryptic,chevin2014niche},
but model assumptions can strongly affect predictions,
including whether variation is due to rare or common alleles \citep{slatkin1994segregation},
and the shape of the fitness landscape \citep{fraisse2016genetics}.
Our main aim is to provide a concrete framework 
that can provide natural predictions of these model parameters
across a general class of models.
Furthermore, tools from system theory allow analytical predictions to be made
for large populations with complex phenotypes
that would be inaccessible to population simulations.

@@

\emph{Note on implementation:}
The \emph{reachable subspace}
% which we denote by $\reachable$,
is defined to be the closure of $\spn(B)$ under applying $A$
(or equivalently, the span of $B, AB, A^2B, \ldots A^{n-1}B$), \revpoint{AE}{6}
% Analogously to this, we define
and the \emph{unobservable subspace}
%$\mathcal{O}$, is the closure of $\spn(C^T)$ under applying $A$.
% denoted $\unobservable$,
is the largest $A$-invariant subspace
contained in the null space of $C$.
% and $\reachable$ is the largest $A$-invariant subspace contained in the image of $B$.)
The four subspaces, $\rno$, $\ro$, $\nrno$, and $\nro$
are defined from these by intersections and orthogonal complements --
$\ro$ refers to the both \emph{reachable and observable} subspace,
while $\nrno$ refers to the \emph{unreachable and unobservable} subspace,
and similarly for $\nro$ and $\rno$.

@@

For the remainder of the paper, we interpret $\allS$ as the neutral set in the fitness landscape, 
along which a large population will drift under environmental and selective stasis. 
This drift need not be purely neutral --
for instance, second-order selection on robustness
will push the species towards ``flatter'' areas of genotype space \citep{rice1998evolution,hermisson2003epistasis}.
Even if the phenotype is constrained and remains constant through evolutionary time, 
the molecular mechanism underpinning it is not constrained and likely will not be conserved.

@@

Exactly how the coefficients 
(i.e., entries of $A$, $B$ and $C$)
of a haploid system inherited by an offspring from a diploid parent
are determined by the parent's two systems
depends on the genetic basis of any variation in the coefficients.
Thanks to the randomness of meiotic segregation,
the result is random to the extent that each parent is heterozygous
for alleles that affect the coefficients.
Since the $i^\text{th}$ row of $A$ summarizes how each gene regulates gene $i$,
and hence is determined by the promoter region of gene $i$,
the elements of a row of $A$ tend to be inherited together,
which will create covariance between entries of the same row.
It is, however, a quite general observation that the variation seen among recombinant systems
is proportional to the difference between the two parental systems.
% This is certainly true if each coefficient is determined by a single nonrecombining locus,
% so that each coefficient in the system produced by meiosis is an independent random choice
% between the two parental coefficients.
% On the other hand, if a coefficient is determined by additive contributions of independently segregating loci,
% then the law of total variance applied after conditioning on the parental origin of each allele
% implies that the variance is equal to one-quarter of the squared difference between parental coefficients.

@@

\begin{figure}[H]
\centering
\includegraphics{figures/conceptual_fig}
\caption{
    \label{fig:conceptual_fig}
    A conceptual figure of the fitness consequences of hybridization:
    axes represent system coefficients (i.e., entries of $A$);
    the line of optimal system coefficients is down in black;
    solid lines give phenotypic distances to the optimum.
    A pair of parental populations are shown in black, along the optimum;
    a hypothetical population of $F_1$s are shown in red,
    and the distribution of $F_2$s is shown in purple.
    The two figures differ in the genetic basis, and hence,
    the distribution of $F_2$ phenotypes:
    \textbf{(left)}
    $F_2$s compose all four mixed homozygotes if variation at both
    traits has a simple, one-locus genetic basis in both populations; and
    \textbf{(right)}
    $F_2$ show a much wider distribution of phenotypes
    if the genetic basis of variation in each population is polygenic.
}
\end{figure}

\subsection*{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
Hybrid phenotypic break down, as a function of genetic distance between
phenotypically equivalent parental oscillators (described in Example \ref{ex:all_osc})
is illustrated in Example \ref{ex:hybrid_osc}.
How quickly do hybrid phenotypes break down as genetic distance between parents increases?
We will quantify how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative weighting function, \revpoint{2}{3}
$h_0(t)$ is the \emph{optimal} impulse response function
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
\label{eqn:distance}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
In practice, we take $\rho(t) = \exp(-t/4\pi)$,
so that fitness is determined by the dynamics of the system over a few multiples of $2 \pi$,
but not longer. \revpoint{AE}{8}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ 
is explicitly worked out in Appendix \ref{apx:H_calc}, and \revpoint{2}{4}
shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry
of the optimal set $\allS$ relative to segregating genetic variation.
% in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set) with respect to the coordinates.

\begin{example}[Hybrid incompatibility: misregulation due to system drift] \label{ex:hybrid_osc}
Offspring of two equivalent systems from Example \ref{ex:all_osc}
can easily fail to oscillate.
For instance, the $F_1$ offspring between homozygous parents at $\tau=0$ and $\tau=-2$
has phenotype $\phi_{F_1}(t) = e^t$, rather than $\phi(t) = \sin t + \cos t$.
However, the coefficients of these two parental systems differ substantially,
probably more than would be observed between diverging populations.
In figure \ref{fig:hybs} we compare the phenotypes for $F_1$ and $F_2$ hybrids between more similar parents,
and see increasingly divergent phenotypes as the difference between the parental systems increases.
(In this example, the coefficients of $A(\epsilon)$ differ from those of $A(0)$ by an average factor of $1+\epsilon/2$;
such small differences could plausibly be caused by changes to promoter sequences.)
This divergence is quantified in Figure \ref{fig:osc_incompat},
which shows that mean distance to optimum phenotype of the $F_1$ and $F_2$ hybrid offspring between $A(0)$ and $A(\epsilon)$
increases with $\epsilon^2$ and $\epsilon$, respectively.
\end{example}

\begin{figure}[H]
  \centering
 % \begin{tabular}{ccc}
 %   \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_hundreth_tau0} &
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_hundrethtau0} \\
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_tenth_tau0.pdf} &
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_tenthtau0.pdf} \\
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F1_half_tau0.pdf} &
 %     \includegraphics[width=0.5\textwidth, height=0.125\paperheight]{examples/osc_F2_halftau0.pdf}
 % \end{tabular}
  %\includegraphics[scale=0.5]{examples/f1f2_newplot}
  \includegraphics{examples/f1f2_small}
  \caption{
    \textbf{(left)} Phenotypes of $F_1$ hybrids between a homozygous $A(0)$ parent and, 
    top-to-bottom, homozygous $A(\sfrac{1}{100})$, $A(\sfrac{1}{10})$, and $A(\sfrac{1}{2})$ parents,
    where $A(\epsilon)$ is defined in figure \ref{fig:all_osc};
    parental coefficients differ by around 0.5\%, 5\%, and 25\% respectively.
    Parental phenotypes ($\sin t + \cos t$) are shown in solid black, and hybrid phenotypes in dashed blue.
    \textbf{(right)} Phenotypes of all $3^4 = 81$ possible $F_2$ hybrids between the same set of parents,
    with parental phenotype again in black.
    %% PLR: There are at most 128 possible F2s, not 256; but many are equivalent; 
    %%   it only matters if the offspring is homo/het/homozygous, so there are 3^4 = 81 possible expression patterns.
    Different colored lines correspond to different $F_2$ hybrids,
    many of which show complete breakdown.
  } \label{fig:hybs}
\end{figure}
